---
title: "NOAA Weather Data Analysis"
author: "Asif Khan"
date: "9/8/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Title:
Data Analysis to show which types of weather events across US are most harmful in terms of population health and which ones have greatest economic consequences

Synopsis:
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.



Data Processing:
Data used in analysis is from U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. 

Downloading and storing data for exploration

```{r datadownload echo = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", cacheOK = TRUE, destfile = "StormData.csv.bz2")

WeatherData <- read.csv("StormData.csv.bz2")

```





